---
title: "Standardised indices to monitor energy droughts"
author: "Sam Allen and Noelia Otero"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standardised indices to monitor energy droughts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body {
text-align: justify}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,comment = "#>", fig.width = 8, fig.height = 6)
```


## Introduction

`SEI` is a package in R to calculate standardized energy indices that can be used to monitor *energy droughts* in renewable power systems. The indices incorporate energy demand and renewable energy production, and constitute analogues of the standardised precipitation index and standardised precipitation evapotranspiration index, two indices that are regularly employed within operational hydro-meteorological drought monitoring systems (Begueria et al. 2014). 

This document introduces the methodology and how to use the `SEI` package for analysing *energy droughts* of wind and solar power generation. 

  
```{r setup, include=FALSE}
library(SEI)
library(dplyr)
```

## Metholody

The indices are straightforward to construct, can be defined on any timescale, and behave in a relative sense, meaning they can readily be compared for regions with different climates and different installed capacities.
To construct the indices, we assume that there exists a time series of previous values of the renewable energy production, $P_{1}, \dots, P_{n}$, and the corresponding residual load, $L_{1}, \dots, L_{n}$. The observations could be on any timescale that is of interest. The general approach to define standardised indices begins by estimating the cumulative distribution function (CDF) corresponding to these previously observed values, which we label $F_{P}$ for the production and $F_{L}$ for the residual load. The estimated CDF is then used to transform the observations onto a standardised scale, exploiting the \textit{probability integral transform} to do so.  

## Use: Case study

The data used for the case study correspond to time series of renewable generation solar and wind for a total of 27 countries across Europe. This data is publicity available at the [University of Reading](https://researchdata.reading.ac.uk/275/). More details about this dataset can be found in Bloomfield et al. 2020. The original data provides the capacity factors for solar and wind, which were later multiplied by the installed capacities (IC). In this case, we use the installed capacities of 2017.

Let's first have a look at the IC.

```{r}
data("data_ic2017")
head(data_ic2017)
```

We can use either country abbreviations or codes to use in the plots. 

```{r}
abbrevs <- get_abbrevs(unique(data_ic2017$country))
s_names <- unlist(lapply(unique(data_ic2017$country), get_shortcountry_names))
```

```{r fig.align = 'center', out.width = '100%'}
plot_ics(data_ic2017, s_names)
```

# Indices calculation

## Raw values

We begin by looking at the raw renewable energy production, in this case, "PWS", which is wind and solar production.
But it can be applied to any other energy indicator, such as, electricity demand or individual renewable energy sources.

```{r}
data("data_supply")
head(data_supply)
```

```{r}
nvars <- c("PWS") 
```


Here, we calculate raw data at hourly and daily. Note that it can calculate at any preferred time scale (e.g., weekly,  two-weekly etc...) and the units are GW.
  
```{r}
raw_1h <- calculate_energyindex_country(data_supply, method = "none", scale = NULL, nvars)
raw_1d <- calculate_energyindex_country(data_supply, method = "none", scale = 24, nvars)
```

```{r}
head(raw_1h$SDEI$Portugal)
```

The output is a data frame that contains the time series of the index. We can visualize the values for one country. As an example we can plot the index for Portugal.

```{r}
country_str <- "Portugal"
variable_str <- "PWS"
year_vec <- 2019
```

Visualize the time series:

```{r fig.width = 7, fig.height = 4}
library(ggplot2)
sdei_raw_reduced <- raw_1h$SDEI[[country_str]] %>% dplyr::filter(type == variable_str)
plot_sdei(sdei_raw_reduced, year_vec,ivar = "SDEI")
```

```{r fig.width = 7, fig.height = 4}
sdei_rawD_reduced <- raw_1d$SDEI[[country_str]] %>% dplyr::filter(type == variable_str)
plot_sdei(sdei_rawD_reduced, year_vec,ivar = "SDEI")
```


## Standarised values

Next, we calculate the standardised renewable energy production index (SREPI) at hourly and daily time scales. Note that the SREPI is dimensionless. 

```{r}
sdei_1h <- calculate_energyindex_country(data_supply, method = "empirical", scale = NULL, nvars)
sdei_1d <- calculate_energyindex_country(data_supply, method = "empirical", scale = 24, nvars)
```


```{r fig.width = 7, fig.height = 4}
sdei_reduced <- sdei_1h$SDEI[[country_str]] %>% dplyr::filter(type == variable_str)
sdeiD_reduced <- sdei_1d$SDEI[[country_str]] %>% dplyr::filter(type == variable_str)
plot_sdei(sdei_reduced, year_vec, ivar = "SDEI")
plot_sdei(sdei_reduced, year_vec, ivar = "SDEI")
```


## Distributions

The corresponding distribution for the raw values:

```{r fig.width = 7, fig.height = 4}
ylabb <- if(variable_str == "res_load_WS"){"RL (GW)"}else if(variable_str == "PWS"){"REP (GW)"}else{""}
plot_sdei_dist(sdei_raw_reduced, n_bins = 50, title = ylabb)
```

and for the standarised distribution:

```{r fig.width = 7, fig.height = 4}
ylabb <- if(variable_str == "res_load_WS"){"SRLI"}else if(variable_str == "PWS"){"SREPI"}else{""}
plot_sdei_dist(sdei_reduced, n_bins = 50, title = ylabb)
```

# Define energy droughts 

These indices can be used to characterize *energy droughts*, which could occur due to low values of the renewable energy production, or high values of the residual load. 

Here, we show how the SEI can be used to characterize *droughts* of renewable energy production (PWS). For that, threshold values are used to categorize the type of *energy production drought*. In this case,  we say that an *energy production drought* occurs if the index falls below -1. The function `def_energy_drought` calculates the main characteristics of droughts: intensity, occurrence, duration and magnitude, in terms of the threshold values previously defined.



```{r}
variable_str <- "PWS"
# Define threshold values to use when defining droughts
threshvals <- c(1, 1.5, 2)*(2*sign(variable_str!="PWS") - 1) 
sdei_1d_pivot <- lapply(sdei_1d$SDEI, function(x) {x%>%dplyr::rename(!!variable_str:=SDEI)})
```

```{r}
sdei_1d_ev    <- lapply(sdei_1d_pivot, def_energy_drought, variable_str, threshvals, higher = variable_str!="PWS")
```

We can now calculate the number of droughts per year in each country. As this analysis only considers 2019, n_year = 1. 

```{r}
n_year <- 1
num_ev <- sapply(sdei_1d_ev, function(x) sum(x$dur != 0)/n_year) 
```

and visualize the frequency of *energy production drought*. Note that this includes all types of *energy production droughts* (i.e., mild, moderate, severe and extreme). See more details in Allen and Otero, 2022.

```{r fig.width = 7, fig.height = 4}
input_data <- data.frame(num_ev, area=s_names)
plot_index_UEmap(input_data, "num_ev", NULL)
```


References
==========

- Beguería S, Vicente-Serrano SM, Reig F, Latorre B. 2014. Standardized precipitation evapotranspiration index (SPEI) revisited: parameter fitting, evapotranspiration models, tools, datasets and drought monitoring. International Journal of Climatology 34(10): 3001-3023.

- Bloomfield, H., Brayshaw, D., and Charlton-Perez, A. (2020). ERA5 derived time series of European country-
aggregate electricity demand, wind power generation and solar power generation: hourly data from 1979-2019.

- Allen S. and Otero N. (2022). Standardised indices to monitor energy droughts.